Deep Strictness: Milestone 2

### 2017-12-03 presentations/report.tex
Reorganize to stash presentation and paper materials in subdirs

\title{Deep Strictness: Milestone 1}
\author{Kenny, Hengchu}

@@

Here are some examples of functions that operate on the \verb|List| data type.
Each of these functions demonstrate different strictness on the input list. For example, the lazy \verb|Context| doesn't inspect the input list at all, and thus is completely lazy. By contrast, \verb|spineStrict| evaluates the structure of the list without forcing its contents to be evaluated. A more complex example of a demand context is \verb|evenStrict| which, like |spineStrict|, evaluates the entire structure of the list, but which also evaluates every other element contained in the list.

@@

\end{document}

>>>>

\title{Deep Strictness: Milestone 2}
\author{Kenny, Hengchu}

@@

Here are some examples of functions that operate on the \verb|List| data type.
Each of these functions demonstrate different strictness on the input list. For
example, the lazy \verb|Context| doesn't inspect the input list at all, and thus
is completely lazy. By contrast, \verb|spineStrict| evaluates the structure of
the list without forcing its contents to be evaluated. A more complex example of
a demand context is \verb|evenStrict| which, like \verb|spineStrict|, evaluates
the entire structure of the list, but which also evaluates every other element
contained in the list.

@@

There is a mechanical transformation between the type of a function and the
corresponding type of its demand specification. Similarly, there is a mechanical
transformation from the type of a value to the type of the representation of the
demand upon it.

In order to perform these transformations, we use a feature in Haskell called
Type Families. Type Families allow us to write recursive functions over the
structure of types to compute other types. The Haskell compiler GHC has another
feature called Generics, which transforms any type to a generic representation
that uses sums, products and arrows. We have named our two Type Families
\verb|Spec| and \verb|Demand| respectively, and here is their definition.

\begin{minted}{haskell}
data Thunk a = T | E a

Demand (a :+: b) = Demand a :+: Demand b
Demand (a :*: b) = Thunk (Demand a) :*: Thunk (Demand b)
Demand (a -> b)  = FuncDemand

Spec (a :+: b)   = Spec a :+: Spec b
Spec (a :*: b)   = Spec a :*: Spec b
Spec (a -> b)    = (a, Spec a) -> Demand b -> Thunk (Demand a)
\end{minted}

We use the data type \verb|Thunk| to introduce an additional layer of
indirection. A value of type \verb|Demand a| can only ever be a prefix of its
corresponding value at the type \verb|a|, and thunks represent the locations
where the remaining structure can be chopped off by a lack of further evaluation
of the data structure.

With higher-order functions, uncurrying all of the arguments of the higher order
function before applying \verb|Spec| is critical to generate the most
fine-grained type of its demand specification. Consider the example \verb|map|
which maps a function over a list.
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
\end{minted}

If we don't uncurry \verb|map|, then its specification will have type
\begin{figure}[H]
  \centering
\begin{minted}{haskell}
(a -> b, (a, Spec a) -> Demand b
                     -> Thunk (Demand a))
-> Demand ([a] -> [b])
-> Thunk (Demand (a -> b))
\end{minted}
\caption{curried \texttt{map} specification type}
\end{figure}

This type just simplifies to
\begin{minted}{haskell}
(a -> b, (a, Spec a) -> Demand b
                     -> Thunk (Demand a))
-> FuncDemand
-> Thunk FuncDemand
\end{minted}

\verb|FuncDemand| only captures whether the function was called or not. However,
in a typical implementation of \verb|map|, the usage of the input function
depends on whether the input list is empty or not. It's impossible to specify
the demand behavior of Haskell standard library's implementation of \verb|map|
using this type signature.

However, consider the uncurried \verb|map|, which has type
\begin{minted}{haskell}
(a -> b, [a]) -> [b]
\end{minted}

\clearpage
Applying \verb|Spec| to this type produces
\begin{figure}[H]
  \centering
\begin{minted}{haskell}
((a -> b, [a]), ((a, Spec a) -> Demand b
                             -> Thunk (Demand a), Spec [a]))
-> Demand [b]
-> Thunk (Demand [a])
\end{minted}
\caption{uncurried \texttt{map} specification type}
\end{figure}

Here, we see that this type in fact takes the input list \verb|[a]| as an input
to the specification as well! This allows a precise and correct specification of
\verb|map|. This observation applies to general higher-order functions.

We will integrate the \verb|Spec| and \verb|Demand| Type Families with GHC's
generics mechanism to automatically derive specification types of algebraic data
types. We will then write generic generators for representations of
specifications and demands, and then we will develop a presentable API for
programmers to write property based test cases to check the demand behaviors of
Haskell programs.
\end{document}
